Grove: A Convergent Collaborative Structure-Editor Calculus

### 2021-03-01 grove-paper.tex
various small edits

@@

\begin{figure}
  \[
    \arraycolsep=0pt
    \begin{array}{lrlll}
      \textrm{Trees:\qquad}  & t & {}\in Tree & {}::={} & \textrm{Con}(v, m) \mid \textrm{Ref}(v) \\
      \textrm{Maps:}         & m & {}\in Map & {}::={} & [\,] \mid (p \mapsto [(\e, t), \ldots, (\e, t)]); m \\
    \end{array}
  \]
  \caption{Syntax of graph decomposition trees as a grammar.}
  \label{fig:Syntax of graph decomposition trees}
\end{figure}

@@

Mapping to b actions

@@

\section{TODO}

\section{Implementation}%
\label{sec:Implementation}
GRV -- how it is implemented, how it connects to the formalism, describe the graph,

@@

Related: hypertree decomposition is an established topic in AI/DB. Apparently,
it's NP-complete in general, so people look for tractable subproblems. Recent
work focuses on reduction to SMT and parallelizability for large db-like
querying of large graphs, but some papers also focus on efficient decompopsition
for "nearly acyclic" graphs. The closest I could find to what we're already
doing are the k-decomp / opt-k-decomp / det-k-decomp algoriths in this paper:
https://dl.acm.org/doi/10.1145/1412228.1412229, except we're getting determinism
for free (i.e., not having to guess-and-check pivot points) by virtue of our
global root and uuid-based total ordering on basically everything.

@@

% \section{Appendix}
% \label{apx:Appendix}

% \subsection{Concepts}

% \begin{tabular}{cl@{\hspace{1.5cm}}cl@{\hspace{1.5cm}}cl@{\hspace{1.5cm}}cl}
%   $\A$ & Actions      & $\S$  & Edge States   \\
%   %  $\C$ & Cursors & $\U$ & UUIDs         \\
%   $\E$ & Edges        & $\V$  & Vertexes      \\
%   $\G$ & Graphs       & $\AA$ & Graph Actions \\
%   $\I$ & Indices      & $\EE$ & Editors       \\
%   $\K$ & Constructors                         \\
% \end{tabular}

% A \emph{UUID} $u \in \U$ is a universally unique identifier.

% \subsubsection{The Target Language}

% \paragraph{Constructor} \par

% A \emph{constructor} $k \in \K$ represents a node in the target language's abstract syntax tree (AST). Let $k_{root}$ be the root AST node.

% \paragraph{Index} \par

% An \emph{index} $i \in \I$ identifies a position, relative to some parent constructor, of a potential child node in the target language's AST. Let $i_{root}$ be the hyper-position of all rooted constructors.

% \subsubsection{The Graph}

% \paragraph{Vertex} \par

% A \emph{vertex} $v = (k, u) \in \V = \K \times \U$ is a distinct instance of constructor $k$; that is, one that can be identified by UUID $u$. Let $v_{root} = (k_{root}, u_{root})$ be the root vertex of a graph. The \emph{rooted vertexes} of a graph are the immediate children of its $v_{root}$.

% \paragraph{Cursor} \par

% %A \emph{cursor} $c = (v,i) \in \C = \V \times \I$ is a reference to the (possibly empty) set of all vertexes with parent vertex $v$ and child index $i$. Let $c_{root} = (v_{root}, i_{root})$ be the \emph{root cursor}, a reference to all of the rooted vertexes, of a graph.

% \paragraph{Edge} \par

% %An \emph{edge} $\e = (c, v, u) \in \E = \C \times \V \times \U$ adds vertex $v$ with the set of vertexes referenced by cursor $c$.

% \paragraph{Edge State} \par

% An \emph{edge state} $s \in \S = \{\bot,+,-\}$ determines whether an edge does not exist ($\bot$), has been created (+), or has been destroyed (-).

% \paragraph{Graph Action} \par

% A \emph{graph action} $A = (\e,s,u) \in \A = \E \times \S \times \U$ is an instance of a binding from edge $\e$ to edge state $s$ that can be identified by UUID $u$. Denote by $[\e :-> s]_{u}$ a binding from edge $\e$ to edge state $s$ (identified by UUID $u$).

% \paragraph{Graph} \par

% A \emph{graph} $G = [\e :-> s]_{u}^{\ast{}} \in \G = \A^{\ast{}}$ is a function from edges to edge states. Denote by $G(\e) = s$ the state $s$ of edge $\e$ in graph $G$, and by $G A$ the extension of graph $G$ by graph action $A$.

% \subsubsection{The Editor}

% \paragraph{Move Action} \par

% %A \emph{move action} $a \in \A^{move} = \{ \Left,\Right,\Up,\Down \} \cup \{ \Select \} \times \C$ re-positions the cursor.

% \paragraph{Edit Action} \par

% An \emph{edit action} $a \in \A^{edit}= \{ \Create \} \times \K \cup \{ \Destroy \} \cup \{ \Restore \} \times \V$ adds or removes an edge at the cursor.

% \paragraph{Communication Action} \par

% A \emph{communication action} $a = (\Send, A^{\ast{}}, u^{\ast{}}) \in \A^{comm} = \{ \Send \} \times \A^{\ast{}} \times \U^{\ast{}}$ applies the graph action sequence $A^{\ast{}}$ to the editors identified by UUIDs $u^{\ast{}}$.

% \paragraph{Action} \par

% An \emph{action} $a \in \A = \A^{move} \cup \A^{edit} \cup \A^{comm}$ describes a change to one or more editors.

% \paragraph{Editor} \par

% %An \emph{editor} $E = (G,c,A_{Q}^{\ast{}},A_{H}^{\ast{}},u) \in \EE = \mathcal{G} \times \C \times \AA^{\ast{}} \times \AA^{\ast{}} \times \U$ is an instance of graph $G$ with cursor $c$, queued graph actions $A_{Q}^{\ast{}}$, and known graph action history $A_{H}^{\ast{}}$, that can be identified by UUID $u$.

% \subsubsection{The Environment} \par

% An \emph{environment} $(E^{\ast{}}, A^{\ast{}}) \in \EE^{\ast{}} \times \AA^{\ast{}}$ is a set of communicating editors $E^{\ast{}}$ that share a global graph action history $A^{\ast{}}$.

% \subsection{Sets and Relations}

% \paragraph{Edges} \par

% \begin{align*}
%   \edges(G)            & = \dom(G)                                       \\
%   \liveEdges(G)        & = \{ \e \in \edges(G) | G(\e)=+ \}              \\
%   \vertexChildren(G,v) & = \{ ((v',i),v'',u) \in \liveEdges(G) | v'=v \} \\
%   \cursorChildren(G,c) & = \{ (c',v,u) \in \liveEdges(G) | c'=c \}       \\
%   \parents(G,v)        & = \{ (c,v',u) \in \liveEdges(G) | v'=v \}
% \end{align*}

% \paragraph{Vertexes} \par

% \begin{align*}
%   \vertexes(G)        & = \{ v \in \V | \exists (c,v,u) \in \edges(G) \}              \\
%   \liveVertexes(G)    & = \{ v \in \V | \exists (c,v,u) \in \liveEdges(G) \}          \\
%   \childVertexes(G,v) & = \{ v' \in \V | \exists (c,v',u) \in \vertexChildren(G,v) \} \\
%   \orphans(G)         & = \{ v \in \liveVertexes(G) | \abs{\parents(G,v)} = 0 \}      \\
%   \multiparents(G)    & = \{ v \in \liveVertexes(G) | \abs{\parents(G,v)} > 1 \}      \\
%   \vertex(G,u)        & = v \in \vertexes(G) \iff (\exists k \in \K)(v = (k,u))
% \end{align*}

% \paragraph{Graph Actions} \par

% $\known(E_1 \cdots E_n) = A_{H1}^{\ast{}} \cdots A_{Hn}^{\ast{}}
%   \iff (\forall j=1,\ldots,n)(E_j = (G_j,c_j,A_{Qj}^{\ast{}},A_{Hj}^{\ast{}},u_j))$

% \section{Operational Semantics}

% \subsection{Movement}

% $\boxed{E,A -->a E,A}$

% \begin{mathpar}
%   \inferrule{
%   \leftIndex(i) = i'
%   }{
%   (G,(v,i),A_Q^{\ast{}},A_H^{\ast{}},u),A \xrightarrow{\Left} (G,(v,i'),A_Q^{\ast{}},A_H^{\ast{}},u),A
%   }

%   \inferrule{
%   \rightIndex(i) = i'
%   }{
%   (G,(v,i),A_Q^{\ast{}},A_H^{\ast{}},u),A \xrightarrow{\Right} (G,(v,i'),A_Q^{\ast{}},A_H^{\ast{}},u),A
%   }

%   \inferrule{
%   \parents(v) = \{(c',v',u')\}
%   }{
%   (G,(v,i),A_Q^{\ast{}},A_H^{\ast{}},u),A \xrightarrow{\Up} (G,c',A_Q^{\ast{}},A_H^{\ast{}},u),A
%   }

%   \inferrule{
%   \cursorChildren(c) = \{(c',(k,u_k),u')\} \\
%   \downIndex(k) = i
%   }{
%   (G,c,A_Q^{\ast{}},A_H^{\ast{}},u),A \xrightarrow{\Down} (G,((k,u_k),i),A_Q^{\ast{}},A_H^{\ast{}},u),A
%   }

%   \inferrule{}{
%   (G,c,A_Q^{\ast{}},A_H^{\ast{}},u),A \xrightarrow{\Select~c'} (G,c',A_Q^{\ast{}},A_H^{\ast{}},u),A
%   }
% \end{mathpar}

% \subsection{Editing}

% \begin{mathpar}
%   \inferrule{
%   \defaultIndex(c) = TODO(Was None) \\
%   u, u_k \in \U \text{ fresh} \\
%   A_H' = [(c,(k,u_k),u) :-> +]
%   }{
%   (G,c,A_Q^{\ast{}},A_H^{\ast{}},u),A \xrightarrow{\Create~k} (G A_H',c,A_Q^{\ast{}},A_H^{\ast{}}A_H',u),A
%   }

%   \inferrule{
%   \defaultIndex(c) = i' \\
%   c' = ((k,u_k),i') \\
%   u, u_k \in \U \text{ fresh} \\
%   \\\\
%   \cursorChildren(G,c) = \{(c_j,v_j,u_j')\}_{j=1}^n \\
%   A_H'^{\ast{}} =
%     [(c  ,(k,u_k),u  ) :-> +]
%     [(c' ,v_j    ,u_j') :-> +]_{j=1}^n
%   [(c_j,v_j    ,u_j') :-> -]_{j=1}^n
%   }{
%   (G,c,A_Q^{\ast{}},A_H^{\ast{}},u),A \xrightarrow{\Create~k}
%   (G A_H'^{\ast{}},c',A_Q^{\ast{}},A_H^{\ast{}}A_H'^{\ast{}},u),A
%   }

%   \inferrule{
%   \cursorChildren(G,c) = \{\e_j\}_{j=1}^n \\
%   A_H'^{\ast{}} = [\e_j :-> -]_{j=1}^n
%   }{
%   (G,c,A_Q^{\ast{}},A_H^{\ast{}},u),A \xrightarrow{\Destroy}
%   (G A_H'^{\ast{}},c,A_Q^{\ast{}},A_H^{\ast{}}A_H'^{\ast{}},u),A
%   }

%   \inferrule{
%   A_H = [(c,v,u') :-> +] \\
%   u' \text{ fresh}
%   }{
%   (G,c,A_Q^{\ast{}},A_H^{\ast{}},u),A \xrightarrow{\Restore~v} (G A_H,c,A_Q^{\ast{}},A_H^{\ast{}}A_H,u),A %
%   }
% \end{mathpar}

% \subsection{Communication}

% \paragraph{Record} \par
% %
% \begin{mathpar}
%   \inferrule{
%   a \in \A^{edit} \\
%   (G,c,A_Q^{\ast{}},A_H^{\ast{}},u),A -->a (G',c',A_Q^{\ast{}},A_H'^{\ast{}},u),A
%   }{
%   (G,c,A_Q^{\ast{}},A_H^{\ast{}},u),A -->a (G',c',A_Q^{\ast{}}a,A_H'^{\ast{}},u),A
%   }
% \end{mathpar}

% \paragraph{Replay} \par

% $\boxed{E,A \xrightarrow{a^{\ast{}}} E,A}$
% %
% \begin{mathpar}
%   \inferrule{
%     E,A -->a E'',A'' \\
%     E'',A'' \xrightarrow{a^{\ast{}}} E',A' \\
%   }{
%     E,A \xrightarrow{aa^{\ast{}}} E',A'
%   }
% \end{mathpar}

% \paragraph{Publish} \par

% $\boxed{(E,A)^{\ast{}} -->a (E,A)^{\ast{}}}$
% %
% \begin{mathpar}
%   \inferrule{
%   E_j = (G_j,c_j,A_{Qj}^{\ast{}},A_{Hj}^{\ast{}},u_j) \\
%   E_j,A_j \xrightarrow{a'^{\ast{}}} (G_j',c_j',A_{Qj}'^{\ast{}}, A_{Hj}'^{\ast{}},u_j),A_j' \\
%   E_j' =
%   ( G_j'
%   , c_j'
%   , A_{Qj}'^{\ast{}} \setminus a'^{\ast{}}
%   , A_{Hj}'^{\ast{}} \setminus \known(E_1 \cdots E_n),u_j) \\
%   j = 1,\ldots,n
%   }{
%   (E_1,A_1) \cdots (E_n,A_n) \xrightarrow{\Send~ a'^{\ast{}} } (E_1',A_1') \cdots (E_n',A_n')
%   }
% \end{mathpar}

% Missing Action.env members: Record / Report / Stop / Replay, Dump / Load, Clone / Drop

% $G \overset{\left(e,s\right)}{\longrightarrow} G\left[ e \mapsto s \sqcup G(e) \right]$

% % Edge.Map.add action.value.edge (join old_state new_state) graph

>>>>

% TODO: make sure we're not claiming to invent the term "unicycular graph"
% TODO: move "algorithmic section" to Implementation section
% TODO: maybe add a forward reference at end here to Implementation section
% TODO: clean up old comments below

@@

% TODO: clean up this figure
% TODO: move to implementation section

\begin{figure}
  \[
    \arraycolsep=0pt
    \begin{array}{lrlll}
      \textrm{Trees:\qquad}  & t & {}\in Tree & {}::={} & \textrm{Vertex}(v, m) \mid \textrm{Ref}(v) \\
      \textrm{Maps:}         & m & {}\in Map & {}::={} & \varnothing \mid p \mapsto [(\e, t), \ldots, (\e, t)]; m \\
    \end{array}
  \]
  \caption{Syntax of graph decomposition trees as a grammar.}
  \label{fig:Syntax of graph decomposition trees}
\end{figure}

@@

% TODO: define a grammar for user actions --> mapping to sequences of underlying graph edits
% TODO: define well-sorted graph & theorem: graph is well sorted => exists a decomposition

move = delete + add

other composite edits

remember commutativity diagram: we want edits on trees to be equivalent to edits on graphs

given a semantics for user actions, we can show that operating on the grove is equivalent to operating on the graph.
This way is more efficient (less annoying?) because then we don't have to deal with decomposition for every edit.
(we probably want to do it this way in Hazel)

OR

We could go the indirect route and avoid having to prove the theorem.
(this way is easier to start with, and it's how GRV works)

% Mapping to b actions

@@

\section{Implementation}%
\label{sec:Implementation}

% TODO: talk about performance: O(decomp) given "size of the graph" (probably linear in # edges)

GRV -- how it is implemented, how it connects to the formalism, describe the graph,

@@

@@
